Profile likelihoods for geostatsistical models using GPU's

### 2022-01-08 pkg/oldgpuRandom/inst/doc/L2021_11/paper2.Rnw
paper2
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@3103 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

<<preliminaries, echo=FALSE, include=TRUE>>=
options(continue="+  ",  prompt="R> ", digits=3, width=66)

opts_chunk$set(fig.height=3,
		fig.width=4.5,dev='png',tidy=TRUE,fig.align='center',background='#FFFFFF00', prompt=TRUE, highlight=FALSE)

@@

\section{Introduction}
some keypoints to mention:\\
1, anisotropic spatial models are important but diffcult to evaluate because it consists of many covariance parameters.\\
2, standard approximations of spatial models (GMRF) don’t allow for anisotropy (at least as implemented). \\
3, gpuBatchMatrix is able to compute the profile LogL for each of the parameters in linear anisotropic spatial models, for dense matrix, use no approximation methods. allow for uncertainty in covariance parameters\\
4, investigated likelihood-based CI’s for variance parameters, wider than using information matrix based methods (gepstatsp)\\

@@

% \begin{theorem}
% \label{R1}
% $\hat{\theta}(Y) \sim \MVN(\theta, I(\theta)^{-1})$.
% \end{theorem}
% %
% \begin{theorem}
% \label{R2}
% $2\{\ell(\hat\theta(Y); Y)-\ell(\theta;Y) \} \sim \chi^2_p ,$ where $p$ is the dimension of $\theta$.
% \end{theorem}
\textit{Geostatsp's methods}\vspace{0.5cm}\\
The \fct{lgm} function in \pkg{geostatsp} package estimates the correlation parameters $\omega=(\phi,\kappa,\nu^2, \varphi, \psi)$ and the transformation parameter $\lambda$ by maximization of the likelihood. Specifically, \fct{lgm} uses the \textit{optim} function to optimize the equation \eqref{profile} (if \fct{reml}=FALSE)  numerically over many different sets of $(\omega, \lambda)$ to find $(\hat{\omega},\hat{\lambda}) = \argmax_{\omega,\lambda} \ell_p(\omega, \lambda; y')$.  \pkg{geostatsp} then treats $\hat{\omega}$ and $\hat{\lambda}$ as fixed known parameters and back substitute them into \eqref{betahat} and \eqref{eq:4} to obtain the MLEs $\hat{\beta}_{\omega,\lambda}$ and $\hat{\sigma}^2_{\omega,\lambda}$.  \pkg{geostatsp} calculates Wald confidence intervals for model parameters in \fct{lgm}. The \fct{numderiv} function inside \fct{lgm} returns the Hessian matrix evaluated at $\hat{\omega}$ and $\hat{\lambda}$, the diagonals of the inverse of the negative Hessian is used for approximating the variance of the parameters.This way does not take into account of uncertainty in correlation parameters. \pkg{geostatsp} does the above computation process in \prog{C}, and (ADD LATER!) which has achieved a great perfomance in terms of speed.

@@

\subsection{CI’s comparison for model parameters}

@@

<<packages, results='hide', message=FALSE>>=
 library(gpuRandom)
 library(gpuR)
 library(data.table)
 #library(geostatsp)
@

% swissrain matplot over betas

%meuse data

\section{A simulation study}

>>>>

<<preliminaries, echo=FALSE, include=TRUE>>=
options(continue="+  ",  prompt="R> ", digits=3, width=66) 
options(scipen=999)
opts_chunk$set(fig.height=3,
		fig.width=4.5,dev='png',tidy=TRUE,fig.align='center',background='#FFFFFF00', prompt=TRUE, highlight=FALSE)

@@

\section{Introduction}

some keypoints to mention:\\
1, anisotropic spatial models are important but diffcult to evaluate because it consists of many covariance parameters.\\
2, standard approximations of spatial models (GMRF) don’t allow for anisotropy (at least as implemented). \\
3, gpuBatchMatrix is able to compute the profile LogL for each of the parameters in linear anisotropic spatial models, for dense matrix, use no approximation methods. allow for uncertainty in covariance parameters\\
4, investigated likelihood-based CI’s for variance parameters, wider than using information matrix based methods (gepstatsp)\\

@@

% \begin{theorem}
% \label{R1}
% $\hat{\theta}(Y) \sim \MVN(\theta, I(\theta)^{-1})$.
% \end{theorem}
% %
% \begin{theorem}
% \label{R2}
% $2\{\ell(\hat\theta(Y); Y)-\ell(\theta;Y) \} \sim \chi^2_p ,$ where $p$ is the dimension of $\theta$.
% \end{theorem}
\textit{Geostatsp's methods}\vspace{0.5cm}\\
The \fct{lgm} function in \pkg{geostatsp} package estimates the correlation parameters $\omega=(\phi,\kappa,\nu^2, \varphi, \psi)$ and the transformation parameter $\lambda$ by maximization of the likelihood. Specifically, \fct{lgm} uses the \textit{optim} function to optimize the equation \eqref{profile} (if \fct{reml}=FALSE)  numerically over many different sets of $(\omega, \lambda)$ to find $(\hat{\omega},\hat{\lambda}) = \argmax_{\omega,\lambda} \ell_p(\omega, \lambda; y')$.  \pkg{geostatsp} then treats $\hat{\omega}$ and $\hat{\lambda}$ as fixed known parameters and back substitute them into \eqref{betahat} and \eqref{eq:4} to obtain the MLEs $\hat{\beta}_{\omega,\lambda}$ and $\hat{\sigma}^2_{\omega,\lambda}$.  \pkg{geostatsp} calculates Wald confidence intervals for model parameters in \fct{lgm}. The \fct{numderiv} function inside \fct{lgm} returns the Hessian matrix evaluated at $\hat{\omega}$ and $\hat{\lambda}$, the diagonals of the inverse of the negative Hessian is used for approximating the variance of the parameters. In this procedure, the uncertainty of the (estimated) correlation parameters is ignored . \pkg{geostatsp} does the above computation process in \prog{C}, and (ADD LATER!) which has achieved a great perfomance in terms of speed.

@@

<<packages, echo=FALSE,results='hide', message=FALSE>>=
 library(gpuRandom)
 library(gpuR)
 library(data.table)
 library(geostatsp)
 library("sp")
library(knitr)
library(kableExtra)
@

\subsection{swissRain example}
<<swissgeostatsp, echo=FALSE, cache=TRUE>>=
data("swissRain")
swissRain$elevation = extract(swissAltitude, swissRain)
swissRes =  lgm( formula=rain~ elevation, 
                     data=swissRain, 
                     grid=20,
                     covariates=swissAltitude, 
                     reml = FALSE,
                     fixBoxcox=FALSE, fixShape=FALSE, fixNugget = FALSE,  #Set to FALSE to estimate the nugget effect parameter.
                     aniso=TRUE )
A<-swissRes$summary[,c('estimate','ci0.1', 'ci0.9')]
@

<<swisssummarycompare, echo=FALSE>>=
table1 <- read.csv("Book1.csv")
colnames(table1) <- c('Parameter','Estimate','95% CI', 'Estimate','95% CI')
knitr::kable(table1, format = "latex", align = c("rcccc"),booktabs = T, linesep = "", caption= "Summary comparison", digits=3)%>%
 kable_styling(full_width = FALSE, position = "center", latex_options = "HOLD_position") %>%
 add_header_above(c(" ", "geostatsp" = 2, "gpuBatchMatrix" = 2)) 
@

<<swiss_profile_many,echo = FALSE, eval=TRUE, cache=TRUE, fig.width=4,fig.height=3, fig.show ="hold", fig.align='center', fig.cap = "profile LogL of slope parameter", out.width = "50%",dev='pdf',fig.pos="H">>=
setwd("/home/ruoyong/diseasemapping/pkg/oldgpuRandom/inst/doc/L2021_11")
knitr::include_graphics(c('profile_logl_swiss.png'))
@

<<swiss_profile_sole_prep, out.width = "100%", fig.align = "center",echo=FALSE>>=
knitr::include_graphics(c('swiss_pro_4.png'))
@

<<range>>=

\subsection{meuse example}

@@

Estimates by \pkg{geostatsp}
<<meusedata, echo=FALSE, cache=TRUE>>=
data(meuse)
coordinates(meuse) <- c("x", "y")
meuse$Y = log(meuse$zinc)
meuse$sqrtdist = sqrt(meuse$dist)

Res1 =  geostatsp::lgm( formula=Y~sqrtdist, 
                        data=meuse, 
                        grid=20,
                        reml = FALSE,
                        fixBoxcox=FALSE, fixShape=FALSE, fixNugget = FALSE,  #Set to FALSE to estimate the nugget effect parameter.
                        aniso=TRUE )
Res1$summary[,c('estimate','ci0.005', 'ci0.995')]
Res1$optim$logL
@

<<meuse_params, cache=TRUE, echo=FALSE>>=
  ParamList = list(
    range = c(exp(seq(log(100), log(173.1643928), len=6)), exp(seq(log(185), log(330), len=8))),
    shape = c(seq(0.1,3.5, len=4), seq(4, 6, len=4)),
    anisoRatio =c( seq(1, 2.301060332, len=3), seq(2.8, 7, len=4)),
    anisoAngleRadians = c( seq(0, 0.274896758, len=4), seq(0.4, 0.8, len=4)),
    nugget = c(seq(0,0.3946587, len=3), seq(0.5, 3, len=5))
  ) 
  params = do.call(expand.grid, ParamList)
@

<<meuse_logl, cache=TRUE, message=FALSE>>=
result3<- gpuRandom::likfitLgmCov(spatialmodel=Res1,
                  params=params, 
                  boxcox=c(1,0,-0.903901505),  
                  paramToEstimate = c('range','shape','nugget','sdNugget', 'anisoRatio','anisoAngleRadians'), 
                  cilevel=0.95, 
                  type = "double",
                  reml=FALSE, 
                  NparamPerIter=400,
                  Nglobal=c(128,64),
                  Nlocal=c(16,16),
                  NlocalCache=2800,
                  verbose=FALSE)
result3$summary
@

\section{A simulation study}
<<simulate_datasets, echo=FALSE, cache=TRUE, results='hide', message=FALSE>>=
############### A SIMULATION STUDY IN THE PAPER 2 ####################################
Nsim=100
set.seed(88)
mydat = SpatialPointsDataFrame(cbind(seq(100*1e3,140*1e3, len=500), seq(100*1e3,120*1e3, len=500)), 
                               data=data.frame(cov1 =stats::rnorm(500, 120, 20), cov2 = stats::rpois(500, 50))
)

## simulate a random field
trueParam = c(range=1000, shape=2, variance=0.5^2, tauSq=1^2, anisoRatio=1, anisoAngleRadians=0)
nugget = trueParam['tauSq']/trueParam['variance']
names(nugget) <- "nugget"
nugget   #4
boxcox = c(boxcox=1)
options("useRandomFields" = FALSE)
for (i in 1:Nsim){
  set.seed(i*6-1)
  U <- geostatsp::RFsimulate(model=trueParam[setdiff(names(trueParam),'tauSq')],x=mydat)@data
  # simulate response
  Y <- 3 + 1*mydat$cov1 + 0.5*mydat$cov2 + U + stats::rnorm(length(mydat), 0, sd=sqrt(trueParam["tauSq"]))
  mydat@data <- cbind(mydat@data,Y)
}
colnames(mydat@data) <- c("cov1", "cov2", paste("Y", c(1:Nsim), sep=""))

## geostatsp's estimates
swissRes =  lgm( formula=Y10~ cov1 + cov2,
                 data=mydat,
                 grid=20,
                 reml = FALSE,
                 boxcox=1,
                 fixBoxcox=TRUE, fixShape=FALSE, fixNugget = FALSE,  #Set to FALSE to estimate the nugget effect parameter.
                 aniso=FALSE )
swissRes$summary[,c('estimate','ci0.005', 'ci0.995')]
swissRes$optim$mle
@

<<ranuparams, echo=FALSE, cache=TRUE>>=
newParamListrn = list(
  range = c(exp(seq(log(0.3), log(1), len=10))*1000, exp(seq(log(1.2), log(2.9), len=12))*1000),  
  nugget = c(seq(0, 4, len=8), seq(4.5, 20, len=12)),  
  shape = 2,
  anisoRatio =1,
  anisoAngleRadians = 0.0
) 
paramsrn = do.call(expand.grid, newParamListrn)
@

<<run_forloop, echo=FALSE, results='hide', message=FALSE, warning=FALSE, cache=TRUE>>=
count = 0
breaksf = rep(0, 10)
makedata <- rep(0, 4)
estimates <- rep(0, 6)
## calculate logl
for (i in 1:Nsim){
  result<-gpuRandom::likfitLgmCov2d(
    data= mydat,
    formula= as.formula(paste(colnames(mydat@data)[i+2], "~", "cov1 + cov2")), 
    coordinates=mydat@coords,
    params=paramsrn, 
    paramToEstimate = c('range','nugget'),
    boxcox = 1,
    cilevel=0.8,
    type = "double",
    reml=FALSE, 
    NparamPerIter=400,
    Nglobal=c(128,64),
    Nlocal=c(16,16),
    NlocalCache=2800,
    verbose=FALSE)

  a <- geostatsp::loglikLgm(
    c(trueParam['range'],
      trueParam['shape'],
      nugget,
      trueParam[c('anisoRatio')],
      trueParam[c('anisoAngleRadians')],
      boxcox),
    data = mydat,
    formula = as.formula(paste(colnames(mydat@data)[i+2], "~", "cov1 + cov2")),
    reml = FALSE,
    minustwotimes=FALSE)[['logLik']]

  if (a  >= result$breaks){
    count = count +1
  }
  breaksf[i] = result$breaks
  makedata0 <- cbind(paramsrn[,1:2], result$LogLik, D=i)
  makedata <- rbind(makedata, makedata0)
  estimates <- cbind(estimates, result$summary)
}
colnames(makedata) <- c("range",'nugget', 'LogLik', 'D')
makedatahey <- makedata[-1,]
estimates_summary <- estimates[,-1]
mean_estimates <- apply(estimates_summary, 1, mean)
@

<<results, cache=TRUE>>=
count/Nsim
mean_estimates
@

<<graphs, echo = FALSE, eval=TRUE, cache=TRUE>>=
library(RColorBrewer)
library('ggplot2')
forigin <- ggplot(subset(makedatahey, D==1), aes(x = range, y = nugget, z=LogLik)) + theme_bw() + 
  scale_y_continuous(limits=c(0,20))+
  scale_x_continuous(limits=c(300,3000) )+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  geom_point(size = 0.5, colour='grey') + 
  stat_contour(breaks=breaksf[1])  + 
  geom_vline(xintercept = trueParam['range'])+
  geom_hline(yintercept=nugget)
cols <- rainbow(Nsim)
for (i in 2:Nsim){
  forigin <- forigin + geom_contour(data = subset(makedatahey, D==i), aes(x = range, y = nugget, z=LogLik), breaks=breaksf[i], col= cols[i])
}
@

<<showgraphs, echo = FALSE, eval=TRUE, cache=TRUE, fig.width=4,fig.height=3, fig.show ="hold", fig.align='center', fig.cap = "range and nugget contours", out.width = "50%",dev='pdf'>>=
#par(mfrow = c(1, 2))
#forigin
knitr::include_graphics(c('p2Rplot_100.png','rsplot.png'))
@
